# Run the analysis in the paper

This document describes the workflow to run the analysis in the paper. The analysis is divided into several steps, including data cleaning, model fitting, and posterior predictive checks. The analysis is performed in R and Stan. The code is available in the `R` folder. The Stan models are available in the `src` folder. The outputs are saved in the `outputs` folder. The figures are saved in the `outputs/figs` folder. 

```{r}

# Load the all the required functions for the analysis 
source("R/main.r")

## Clean the NIH data
clean_nih_data("wu_s")
clean_nih_data("wu_rbd")

## Clean the SID data
clean_sid_data( "wu_s")
clean_sid_data( "wu_rbd")

# Plot the cleaned data
plot_data_clean()

# Make table 1
make_table1()

# Make prosterior predicrive checks
plot_prior_pred()
create_find_cvalue()

```

# Run HMC sampler
 
The following code runs the HMC sampler in Stan. The code is available in the `R` folder. The Stan models are available in the `src` folder. 

```{r}

# Run HMC sampler in stan
run_sim <- function(name, model, data_list, folder_name) {
    fit_ode_model <- model$sample(
        data = data_list,    # named list of data
        refresh = 100,
        iter_warmup = 1000,          # number of warmup iterations per chain
        iter_sampling = 1000,            # total number of iterations per chain
        chains = 4,            # number of cores (could use one per chain)
        parallel_chains = 4,
        adapt_delta = 0.95,
        max_treedepth = 10,
        init = 0
    )
    fit_ode_model$save_object(file = here::here("outputs", "stanfit", paste0(name, ".RDS")))
}

# Working for NIH

ode_model <- cmdstan_model(here::here("src", "cauchy_rk45_waneLLPC_ind.stan"), compile = TRUE)
ode_model_m3_ind <- ode_model$compile(stanc_options = list("O1", "canonicalize"))

data_list_nih_s <- readRDS(here::here("outputs", "data_clean", "nih_wu_s", "list_stan_data.RDS"))
data_list_nih_rbd <- readRDS(here::here("outputs", "data_clean", "nih_wu_rbd", "list_stan_data.RDS"))

run_sim("nih_vac_wu_s_i", ode_model_m3_ind, data_list_nih_s)
run_sim("nih_vac_wu_rbd_i", ode_model_m3_ind, data_list_nih_rbd)

```

#  Inspect model fit
 
The following code inspects the model fit for both the spike and RBD antigen. Main reason is to find the source of divergence. Plots are are in the `outputs/figs/diag` folder.

```{r}

# Inspect divergent transitions and treedepth

source(here::here("R", "met_mcmc.R"))

stanfit_name <- "nih_vac_wu_s_i"
fit_model <- readRDS(file = here::here("outputs", "stanfit",  paste0(stanfit_name, ".RDS")) )
  

param_vec_A <- c("param1", "param3", "param4", "sigmaa1", "sigmaa3", "sigmaa4")
param_vec_B <- c("param1", "param3", "param4", "sigmat1", "sigmat3", "sigmat4")
param_vec_C <- c("param1", "param3", "param4", "sigmav1", "sigmav3", "sigmav4")
param_vec_D <- c("param1", "param3", "param4", "sigmai1", "sigmai3", "sigmai4")

run_sampler_diag(fit_model)
plot_diag(fit_model, param_vec_A, "A")
plot_diag(fit_model, param_vec_B, "B")
plot_diag(fit_model, param_vec_C, "C")
plot_diag(fit_model, param_vec_D, "D")


```


# Inspect convergence and autocorrelation
  
The following code inspects the convergence and autocorrelation of the model fit for both the spike and RBD antigen. Plots are are in the `outputs/figs/cc` folder.

```{r}

source(here::here("R", "met_mcmc.R"))

stanfit_name <- "nih_vac_wu_rbd_i"
fit_model <- readRDS(file = here::here("outputs", "stanfit",  paste0(stanfit_name, ".RDS")) )
sampler_diagnostics <- fit_model$sampler_diagnostics()
df_sampler_diagnostics <- as_draws_df(sampler_diagnostics) 

run_sampler_diag(fit_model)

run_cc_antigen( "nih_vac_wu_s_i", "s")
run_cc_antigen( "nih_vac_wu_rbd_i", "rbd")

```

# Get posterior trajectories

The following code gets the posterior trajectories for both the spike and RBD antigen. The code is available in the `R` folder. The Stan models are available in the `src` folder. The outputs are saved in the `outputs` folder.
These can take a while to run! But you can plot the figures without running them. 

```{r}
source(here::here("R", "utils.R"))
#source(here::here("R", ".R"))

stanfit_name <- "nih_vac_wu_s_i"
foldername <- "nih_wu_s"
stanfit_s <- readRDS(file = here::here("outputs", "stanfit", paste0(stanfit_name, ".RDS")) )
stanfit <- stanfit_s
stanfit_name <- "nih_vac_wu_rbd_i"
stanfit_rbd <- readRDS(file = here::here("outputs", "stanfit", paste0(stanfit_name, ".RDS")) )

run_full_posterior_tracjectories(stanfit_s, "nih_wu_s")
run_full_posterior_tracjectories(stanfit_rbd, "nih_wu_rbd")

run_val_trajectories(stanfit_s, "sid_wu_s")
run_val_trajectories(stanfit_rbd, "sid_wu_rbd")

```

# Plot the posterior trajectories of the calibration and training dataset

The following code plots the posterior trajectories of the calibration and training dataset for both the spike and RBD antigen. 
This is essentially Figure 1.

```{r}

source(here::here("R", "res_fit_cal.R"))
source(here::here("R", "res_fit_val.R"))

plot_cal_traj_pop()
plot_cal_crps_ind()
plot_cal_traj_ind()

plot_val_traj_pop()
plot_val_crps_ind()
plot_val_traj_ind()

```

```{r}

plot_posteriors(stanfit_s, "nih_wu_s")
plot_marg_posteriors(stanfit_s, "nih_wu_s")
plot_posteriors_decay(stanfit_s, "nih_wu_s")

plot_posteriors(stanfit_rbd, "nih_wu_rbd")
plot_marg_posteriors(stanfit_rbd, "nih_wu_rbd")
plot_posteriors_decay(stanfit_rbd, "nih_wu_rbd")

```

```{r}

plot_bcell(stanfit_s, "nih_wu_s")
plot_bcell(stanfit_rbd, "nih_wu_rbd")

# Stuff for manuscript 


recode_expo <- c("2" = "ChAdOx1", "1" = "BNT162b2")

relabel_model_state <- c("1" = "a1: Rate of B-cell proliferation",
    "2" = "a2: Rate of derivation to plasmablasts", 
    "3" = "a3: Rate of neutralising antibody production from plasmablasts", 
    "4" = "a4: Rate of neutralising antibody production from plasma cells")
comparison_t <- stanfit_rbd$draws(c("theta_t")) %>%
    spread_draws(theta_t[i, v, s]) %>% 
        mutate(s = recode(s, !!!relabel_model_state)) %>%
        mutate(v = recode(v, !!!recode_expo)) %>%
        filter(s != "a2: Rate of derivation to plasmablasts") %>% 
        pivot_wider(names_from = s, values_from = theta_t) 

comparison_t_mean <- comparison_t %>% group_by(i) %>% summarise(
        `a1: Rate of B-cell proliferation` = mean_qi(`a1: Rate of B-cell proliferation`),
        `a3: Rate of neutralising antibody production from plasmablasts` = mean_qi(`a3: Rate of neutralising antibody production from plasmablasts`),
        `a4: Rate of neutralising antibody production from plasma cells` = mean_qi(`a4: Rate of neutralising antibody production from plasma cells`))#%>%
      #  mutate(i = recode(i, "1" = "<28 days", "2" = "\u2265 28 days")) 



```

# Plot the predicted antibody kinetics

```{r}

plot_ab_kinetics(stanfit_s, "nih_wu_s")
plot_ab_kinetics(stanfit_rbd, "nih_wu_rbd")

plot_abkin_tt()

```


```{r}


relabel_model_vaccine <- c("2" = "ChAdOx1", "1" = "BNT162b2")
relabel_l_s <-c("1" = "Memory B-cell conc.",
    "2" = "Plasmablast conc.", 
    "3" = "sVNT from plasmablasts", "4" = "GC", "5" = "PB", 
    "6" = "sVNT from plasma cells", "tot" = "sVNT total")

full_dynamics_vac_2_wu_s <- stanfit_s$draws(c("y_full_vac")) %>%
        spread_draws(y_full_vac[i, t, s])
full_dynamics_vac_2a_wu_s <- full_dynamics_vac_2_wu_s %>% filter(s %in% c(1, 2, 3, 6)) %>%
    pivot_wider(names_from = "s", values_from = "y_full_vac") %>%
    mutate(tot = `3` + `6`, prop = `3` / (`3` + `6`)) %>% 
    pivot_longer(`1`:`prop`, names_to = "s", values_to = "y_full_vac") %>%
    group_by(i, t, s) %>%
    mean_qi(y_full_vac) 

tt_df <- full_dynamics_vac_2a_wu_s %>% mutate(s = recode(s, !!!relabel_l_s)) %>%
    filter(s == "prop") %>%
    mutate(vax_type = recode(i, !!!relabel_model_vaccine)) %>% drop_na %>% filter(y_full_vac < 0.5) %>% 
    group_by(vax_type) %>% filter(t == min(t)) 

p1 <- full_dynamics_vac_2a_wu_s %>% mutate(s = recode(s, !!!relabel_l_s)) %>%
    filter(s != "prop") %>%
    filter(!s %in% c("Titre (short-lived) (log)", "Titre (long-lived) (log)")) %>% 
    mutate(vax_type = recode(i, !!!relabel_model_vaccine)) %>% drop_na %>% 
    filter(s %in% c(
        "sVNT from plasmablasts",
        "sVNT from plasma cells", 
        "sVNT total")) 

p1 %>% filter(s == "sVNT total", vax_type == "ChAdOx1") %>% filter(y_full_vac == max(y_full_vac))
p1 %>% filter(s == "sVNT total", vax_type == "ChAdOx1") %>% filter(t == 250)

p1 %>% filter(s == "sVNT total", vax_type == "ChAdOx1") %>% filter(y_full_vac < log10(10))


p1 %>% filter(s == "sVNT total", vax_type == "BNT162b2") %>% filter(y_full_vac == max(y_full_vac))
p1 %>% filter(s == "sVNT total", vax_type == "BNT162b2") %>% filter(t == 250)
p1 %>% filter(s == "sVNT total", vax_type == "BNT162b2") %>% filter(y_full_vac < log10(10))

df_uncert_all_wu_t <- clean_fig4_alt(stanfit_s, "nih_wu_s")
df_uncert_all_wu_t_2 <- df_uncert_all_wu_t %>% mutate(v = recode(v, "2" = "ChAdOx1", "1" = "BNT162b2")) 
p  <- ggplot(df_uncert_all_wu_t_2, aes(time)) + stat_ecdf() + facet_grid(vars(v))
pg <- ggplot_build(p)$data[[1]] %>% rename(v = PANEL) %>% filter(x >= 0, x < 365) %>%
    select(y, x, v) %>% mutate(y = 1-y) %>%  mutate(v = recode(v, "2" = "ChAdOx1", "1" = "BNT162b2")) 
to_add_max <- pg %>% group_by(v) %>% summarise(y = max(y)) %>% mutate(x = 0)
to_add_min <- pg %>% group_by(v) %>% summarise(y = min(y)) %>% mutate(x = 365)
pg <- bind_rows(pg, to_add_min) %>% bind_rows(to_add_max)
pg %>% filter(v == "BNT162b2") %>% arrange(y)
pg %>% filter(v == "ChAdOx1") %>% arrange(y)

df_uncert_all_wu_t <- clean_fig4_alt_time(stanfit_s, "nih_wu_s")
df_uncert_all_wu_t_2 <- df_uncert_all_wu_t %>% convert_to_data_time
df_uncert_all_wu_t_2 %>% filter(v == "BNT162b2", time_d1 == "<28 days") %>% arrange(y)
df_uncert_all_wu_t_2 %>% filter(v == "BNT162b2", time_d1 == ">= 28 days") %>% arrange(y)

df_uncert_all_wu_t_2 %>% filter(v == "ChAdOx1", time_d1 == "<28 days") %>% arrange(y)
df_uncert_all_wu_t_2 %>% filter(v == "ChAdOx1", time_d1 == ">= 28 days") %>% arrange(y)


```


# Plot/calcualte the predictive power on validation data

```{r}

compare_cprs_antibodies( stanfit_s, "sid_wu_s") 
compare_cprs_antibodies( stanfit_rbd,  "sid_wu_rbd") 

prediction_assess(stanfit_s, "sid_wu_s")
prediction_assess(stanfit_rbd, "sid_wu_rbd")

prediction_assess_compare()

```